Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-08-28 paper/supplemental_information.Rmd
add creel re-analysis to supplement

---
title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
  - \renewcommand{\thefigure}{S\arabic{figure}}
  - \renewcommand{\thetable}{S\arabic{table}}

---

@@

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{llllllllll}
\toprule
& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{RT} & \multicolumn{1}{l}{log RT} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
\midrule
age &  1.00 &  &  &  &  &  &  & \\
log age  & 0.98 & 1.00 &  &  &  &  &  &\\
rt  & -0.33 & -0.35 & 1.00 &  &  &  &  & \\
log rt  & -0.34 & -0.36 & 0.96 & 1.00 &  &  &  & \\
long window accuracy  & 0.44 & 0.48 & -0.48 & -0.46 & 1.00 &  &  & \\
short window accuracy & 0.38 & 0.43 & -0.62 & -0.61 & 0.82 & 1.00 &  & \\
production vocabulary & 0.72 & 0.70 & -0.31 & -0.33 & 0.51 & 0.45 & 1.00 & \\
comprehension vocabulary & 0.42 & 0.42 & -0.25 & -0.24 & 0.24 & 0.24 & 0.59 & 1.00\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
\end{table}

```{r corrs, results = "asis", include = FALSE}

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & Long window, linear age & 0.14\\
48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & Long window, log age & 0.12\\
50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & Short window, linear age & 0.09\\
50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & Short window, log age & 0.08\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between accuracy and age. \label{tab:accmods}}
\end{table}

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & Log RT, linear age & 0.21\\
18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & Log RT, log age & 0.21\\
18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & Linear RT, linear age & 0.18\\
18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & Linear RT, log age & 0.18\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between RT and age. \label{tab:rtmods}}
\end{table}

```{r accmods, results="asis", include = FALSE}
papaja::apa_table(acc_mods_lmer_summary)
```

@@

In the literature on the "law of practice", although the log-log relationship we observed is commonly present in the aggregate across individuals, the situation is substantially more complex when relationships are measured within individuals. The best fitting curves for individuals are often exponentials or delayed exponentials  (Evans et al., 2018; Heathcote, Brown, \& Mewhort, 2000). With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time. Thus, our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.

@@

<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
<!-- % Mon Apr 14 12:25:14 2025 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
\hline
& F1 & F2 & F3 \\ 
\hline
RT & -0.19 & 0.81 & -0.30 \\ 
RT var & -0.10 & 0.81 & -0.22 \\ 
long window accuracy & 0.33 & -0.31 & 0.55 \\ 
long window accuracy var & -0.10 & 0.26 & -0.65 \\ 
production vocabulary & 0.95 & -0.04 & 0.30 \\ 
comprehension vocabulary & 0.61 & -0.16 & -0.01 \\ 
age & 0.63 & -0.14 & 0.37 \\ 
\hline
\end{tabular}
\caption{\label{tab:loadings}Factor loadings for the exploratory three factor solution using varimax rotation.} 
\end{table}

# Factor Analysis: First Administrations

@@

* a one-factor model;
* a two-factor model with vocabulary separated from speed and accuracy;
* a two-factor model with speed separated from accuracy and vocabulary; and 
* a two-factor model with variability terms separated from speed, accuracy, and vocabulary.

@@

xtable(anova(fa1_fit, fa2_fit, fa2b_fit, fa2v_fit, fa3_fit), digits = 2)
```
<!-- xtable 1.8-4 package -->
<!-- % Fri Apr  4 15:15:52 2025 -->
\begin{table}[ht]
\centering
\caption{Model comparison for alternative factor structures. $p$-values show differences between adjacent models; no $p$-values are shown for comparisons between non-nested models. \label{tab:modcomp}}
\begin{tabular}{lrrrrrrrr}
\hline
& Df & AIC & BIC & Chisq & Chisq diff & RMSEA & Df diff & Pr($>$Chisq) \\ 
\hline
Three-factor  & 6 & 46346.69 & 46475.43 & 91.96 &  &  &  &  \\ 
Two-factor (vocab)  & 8 & 46397.74 & 46514.22 & 147.01 & 55.05 & 0.09 & 2 & $<$ 0.0001 \\ 
Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0 &  \\ 
One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 
\hline
\end{tabular}
\end{table}

# Non-linear Growth Models

To test for the differentiation of vocabulary growth based on initial reaction time, we used the package `nlme` to fit a logistic growth model to the production data. This model has two parameters for the logistic curve, a scale and an intercept. Both were allowed to interact with initial reaction time. We also included random effects of logistic intercept and scale by participant and a grouping term across datasets. This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}). 
```{r}
# xtable(summary(mod_nlme)$tTable)
```

<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->
<!-- % Thu Mar 13 13:10:30 2025 -->
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
\hline
& Value & Std Error & DF & $t$-value & $p$-value \\ 
\hline
Growth Intercept & 2.26 & 2.70 & 1766.00 & 0.84 & 0.40 \\ 
Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & < 0.0001 \\ 
Growth Scale & 2.05 & 2.60 & 1766.00 & 0.79 & 0.43 \\ 
Growth Scale $\times$ $t_0$ RT & 0.28 & 0.38 & 1766.00 & 0.75 & 0.46 \\ 
\hline
\end{tabular}
\caption{Fixed effects estimates from logistic growth model. \label{tab:nlme}}
\end{table}

Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling. Thus, a shift in quadratic growth in the linear model and a shift in intercept in the logistic model both point to the same overall effect, which is faster growth at the point of maximal sensitivity of the CDI. Neither model can estimate whether the overall growth trajectory is different beyond the range of the CDI. Thus, although these models might initially seem to be in conflict, we believe that they actually point to the same phenomenon, which is perhaps better described by the longitudinal SEM model reported in the main text. Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.

@@

Heathcote, A., Brown, S., \& Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.

>>>>

---
title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
  - \renewcommand{\thefigure}{S\arabic{figure}}
  - \renewcommand{\thetable}{S\arabic{table}}
---

@@

# Comparison of Reaction Times for Correct and Incorrect Trials: Re-analysis of Creel (2024)

```{r}
#load data using Creel (2024) analysis scripts
# To run this script, you need to first pull in two datasets from the Creel (2024) OSF repository (https://osf.io/wszd5/):
#allfamiliars.csv and allnovels.csv
library(osfr)
creel_proj <- osf_retrieve_node("wszd5")
#donwload familiar trials
#TO DO: separate this out so files are downloaded only once
creel_proj %>%
  osf_ls_files(pattern = "allfamiliars.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")
creel_proj %>%
  osf_ls_files(pattern = "allnovels.csv") %>% # Replace "data.csv" with your file's name
  osf_download(path = here("supplemental_data","creel_2024"),conflicts = "overwrite")

#run Creel prep script
source(here("supplemental_data","creel_2024","Step2aCorrsNDataPrep.R"))
# run Peekbank-style rt computation over Creel data
source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))

 creel_plot <- ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
  geom_smooth(method="lm")+
  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
  theme_cowplot(font_size=20)+
  #add correlation and p-value as text box
  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=9)+
  theme(legend.position=c(0.8,0.2))+
  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
ggsave(here("supplemental_data","creel_2024","creel_rt_correlation_correcttrials_vs_alltrials.png"),width=9,height=6)

creel_correlation <- cor.test(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs")

The Peekbank dataset only includes measurements of infants' looking behavior, with no measure of a final target selection. This contrasts with work in the visual-world paradigm with older children and adults, in which participants make a final explicit choice about which image matches the target label (e.g. Colby & McMurray, 2023). Having this additional response allows a clearer separation of accuracy and reaction times, because researchers can compute reaction times specifically on those trials in which participants responded correctly. This strategy helps avoid a possible mixing of reaction times for incorrect and correct responses, which might be generated by different underlying cognitive processes. A possible concern with the Peekbank datasets --- and reaction times in infant looking-while-listening studies more generally --- is that it is difficult to separate reaction times for correct vs. incorrect responses in the absence of an independent final choice response.

To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024). This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image. Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e. focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent. The results are visualized below. Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`). This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children. While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.

```{r}
creel_plot
```

@@

```{=tex}
\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{llllllllll}
\toprule
& \multicolumn{1}{l}{age} & \multicolumn{1}{l}{log age} & \multicolumn{1}{l}{RT} & \multicolumn{1}{l}{log RT} & \multicolumn{1}{l}{long acc} & \multicolumn{1}{l}{short acc } & \multicolumn{1}{l}{prod} & \multicolumn{1}{l}{comp}\\
\midrule
age &  1.00 &  &  &  &  &  &  & \\
log age  & 0.98 & 1.00 &  &  &  &  &  &\\
rt  & -0.33 & -0.35 & 1.00 &  &  &  &  & \\
log rt  & -0.34 & -0.36 & 0.96 & 1.00 &  &  &  & \\
long window accuracy  & 0.44 & 0.48 & -0.48 & -0.46 & 1.00 &  &  & \\
short window accuracy & 0.38 & 0.43 & -0.62 & -0.61 & 0.82 & 1.00 &  & \\
production vocabulary & 0.72 & 0.70 & -0.31 & -0.33 & 0.51 & 0.45 & 1.00 & \\
comprehension vocabulary & 0.42 & 0.42 & -0.25 & -0.24 & 0.24 & 0.24 & 0.59 & 1.00\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Pairwise correlations between primary variables of interest. \label{tab:corrs}}
\end{table}
```
```{r corrs, results = "asis", include = FALSE}

@@

```{=tex}
\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
48354 & 0.28 & -8,338.83 & 16,691.65 & 16,753.16 & 16,677.65 & 48347 & Long window, linear age & 0.14\\
48354 & 0.28 & -8,305.39 & 16,624.78 & 16,686.28 & 16,610.78 & 48347 & Long window, log age & 0.12\\
50244 & 0.31 & -12,486.60 & 24,987.21 & 25,048.98 & 24,973.21 & 50237 & Short window, linear age & 0.09\\
50244 & 0.31 & -12,458.77 & 24,931.54 & 24,993.31 & 24,917.54 & 50237 & Short window, log age & 0.08\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between accuracy and age. \label{tab:accmods}}
\end{table}
```
```{=tex}
\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\begin{tabular}{lllllllll}
\toprule
n. obs & \multicolumn{1}{c}{sigma} & \multicolumn{1}{c}{logLik} & \multicolumn{1}{c}{AIC} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{REMLcrit} & \multicolumn{1}{c}{df.residual} & \multicolumn{1}{c}{model} & \multicolumn{1}{c}{$r_2$}\\
\midrule
18940 & 0.45 & -12,414.69 & 24,843.38 & 24,898.32 & 24,829.38 & 18933 & Log RT, linear age & 0.21\\
18940 & 0.45 & -12,393.27 & 24,800.53 & 24,855.48 & 24,786.53 & 18933 & Log RT, log age & 0.21\\
18940 & 570.49 & -147,644.43 & 295,302.87 & 295,357.81 & 295,288.87 & 18933 & Linear RT, linear age & 0.18\\
18940 & 570.02 & -147,622.67 & 295,259.34 & 295,314.29 & 295,245.34 & 18933 & Linear RT, log age & 0.18\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\caption{Model comparison metrics for different functional forms of the relationship between RT and age. \label{tab:rtmods}}
\end{table}
```
```{r accmods, results="asis", include = FALSE}
papaja::apa_table(acc_mods_lmer_summary)
```

@@

In the literature on the "law of practice", although the log-log relationship we observed is commonly present in the aggregate across individuals, the situation is substantially more complex when relationships are measured within individuals. The best fitting curves for individuals are often exponentials or delayed exponentials (Evans et al., 2018; Heathcote, Brown, & Mewhort, 2000). With our current dataset we cannot specifically determine whether within-individual patterns of change conform to linear, power law, or exponential developmental patterns unfortunately, because we have insufficient data about individuals' improvement across time. Thus, our current results apply to the form of the age gradient as opposed to the form of any individual's pattern of developmental change.

@@

<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->

<!-- % Mon Apr 14 12:25:14 2025 -->

```{=tex}
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
\hline
& F1 & F2 & F3 \\ 
\hline
RT & -0.19 & 0.81 & -0.30 \\ 
RT var & -0.10 & 0.81 & -0.22 \\ 
long window accuracy & 0.33 & -0.31 & 0.55 \\ 
long window accuracy var & -0.10 & 0.26 & -0.65 \\ 
production vocabulary & 0.95 & -0.04 & 0.30 \\ 
comprehension vocabulary & 0.61 & -0.16 & -0.01 \\ 
age & 0.63 & -0.14 & 0.37 \\ 
\hline
\end{tabular}
\caption{\label{tab:loadings}Factor loadings for the exploratory three factor solution using varimax rotation.} 
\end{table}
```
# Factor Analysis: First Administrations

@@

-   a one-factor model;
-   a two-factor model with vocabulary separated from speed and accuracy;
-   a two-factor model with speed separated from accuracy and vocabulary; and
-   a two-factor model with variability terms separated from speed, accuracy, and vocabulary.

@@

xtable(anova(fa1_fit, fa2_fit, fa2b_fit, fa2v_fit, fa3_fit), digits = 2)
```

<!-- xtable 1.8-4 package -->

<!-- % Fri Apr  4 15:15:52 2025 -->

```{=tex}
\begin{table}[ht]
\centering
\caption{Model comparison for alternative factor structures. $p$-values show differences between adjacent models; no $p$-values are shown for comparisons between non-nested models. \label{tab:modcomp}}
\begin{tabular}{lrrrrrrrr}
\hline
& Df & AIC & BIC & Chisq & Chisq diff & RMSEA & Df diff & Pr($>$Chisq) \\ 
\hline
Three-factor  & 6 & 46346.69 & 46475.43 & 91.96 &  &  &  &  \\ 
Two-factor (vocab)  & 8 & 46397.74 & 46514.22 & 147.01 & 55.05 & 0.09 & 2 & $<$ 0.0001 \\ 
Two-factor (speed) & 8 & 46486.96 & 46603.44 & 236.23 & 89.22 & 0.00 & 0 &  \\ 
Two-factor (variability) & 8 & 46536.10 & 46652.58 & 285.37 & 49.14 & 0 & 0 &  \\ 
One-factor & 9 & 46535.49 & 46645.84 & 286.76 & 1.39 & 0.01 & 1 & 0.24 \\ 
\hline
\end{tabular}
\end{table}
```
# Non-linear Growth Models

To test for the differentiation of vocabulary growth based on initial reaction time, we used the package `nlme` to fit a logistic growth model to the production data. This model has two parameters for the logistic curve, a scale and an intercept. Both were allowed to interact with initial reaction time. We also included random effects of logistic intercept and scale by participant and a grouping term across datasets. This model showed a significant effect of initial reaction time on the intercept of the logistic growth curve, but not on its scale (see Table \ref{tab:nlme}).

```{r}
# xtable(summary(mod_nlme)$tTable)
```

<!-- % latex table generated in R 4.4.2 by xtable 1.8-4 package -->

<!-- % Thu Mar 13 13:10:30 2025 -->

```{=tex}
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
\hline
& Value & Std Error & DF & $t$-value & $p$-value \\ 
\hline
Growth Intercept & 2.26 & 2.70 & 1766.00 & 0.84 & 0.40 \\ 
Growth Intercept $\times$ $t_0$ RT & 3.30 & 0.38 & 1766.00 & 8.59 & < 0.0001 \\ 
Growth Scale & 2.05 & 2.60 & 1766.00 & 0.79 & 0.43 \\ 
Growth Scale $\times$ $t_0$ RT & 0.28 & 0.38 & 1766.00 & 0.75 & 0.46 \\ 
\hline
\end{tabular}
\caption{Fixed effects estimates from logistic growth model. \label{tab:nlme}}
\end{table}
```
Interpretation of growth in both this model and the linear growth model in the main text is complicated by the fact that the CDI form puts a ceiling on the total number of words that can be recorded; both the quadratic growth functions and the logistic functions come together at the form ceiling. Thus, a shift in quadratic growth in the linear model and a shift in intercept in the logistic model both point to the same overall effect, which is faster growth at the point of maximal sensitivity of the CDI. Neither model can estimate whether the overall growth trajectory is different beyond the range of the CDI. Thus, although these models might initially seem to be in conflict, we believe that they actually point to the same phenomenon, which is perhaps better described by the longitudinal SEM model reported in the main text. Children with greater skill in word recognition show an overall positive shift in the growth trajectory of vocabulary development.

@@

Heathcote, A., Brown, S., & Mewhort, D. J. K. (2000). The power law repealed: The case for an exponential law of practice. Psychonomic Bulletin & Review, 7(2), 185–207.

Creel, S. (2024). Connecting the tots: Strong looking-pointing correlations in preschoolers' word learning and implications for continuity in language development. Child Development, 96, 87-103.
